Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-02 paper.tex
Another pass at intro narrative

%% NOTE: this is much too long currently, but keeping for now so we
% can see which stuff migrates to the intro
\begin{abstract}
\textbf{Background:}
Variant Call Format (VCF) is the standard file format for interchanging
genetic variation data and associated quality control metrics.
It provides a well-defined data model and is central to a large ecosystem
of tools interoperating through this format.
The standard row-based encoding of the VCF data model (either as text
or packed binary) emphasises efficient retrieval of all data for a given
variant, but accessing subsets of the data within rows is inefficient.
Biobank-scale datasets currently available 
consist of hundreds of thousands of whole genomes with tens of terabytes
of compressed VCF per chromosome.
Row-based data storage is fundamentally unsuited to such datasets
and these large VCFs represent a major computational bottleneck.

@@

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, storing information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank-scale projects such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
While there are many advantages to encoding data in text files,
VCF is simply not fit for purpose as part of computational workflows 
at this scale. 
Methods that require efficient access to data
either require VCF data to be converted into 
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
format~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use 
custom binary formats~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate} 
that support the required access patterns.
While PLINK and BGEN support efficient access to genotype data, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy. PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format also leads to large files, particularly for recent
whole-genome sequence datasets. For example, at two bits 
per genotype, the full matrix for the GraphTyper SNP dataset in the UKB
WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

The main solution that has emerged to these problems of 
computation on genetic variation data at scale is Hail~\cite{ganna2016ultra,hail2024}.
Building on components from the Hadoop ecosytem~\citep{white2012hadoop},
Hail distributes computations transparently over a cluster, enabling
interactive analysis of very large datasets. It has been successfully used 
in many studies, and instrumental in generating datasets such as 
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
While Hail is powerful software, it is not a suitable platform for 
developing inference or performing complex analysis. It is complex,
vertically integrated software, with an undocumented file format
[say this better]. 
Similarly, commerical solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~[Terra? BigQuery?]
or single-vendor technologies~\cite[e.g.][]{tiledb2024}.

>>>>

%% NOTE: this is much too long currently, but keeping for now so we
% can see which stuff migrates to the intro
\begin{abstract}
\textbf{Background:}
Variant Call Format (VCF) is the standard file format for interchanging
genetic variation data and associated quality control metrics.
It provides a well-defined data model and is central to a large ecosystem
of interoperating tools.
The standard row-wise encoding of the VCF data model (either as text
or packed binary) emphasises efficient retrieval of all data for a given
variant, but accessing subsets of the data within rows is inefficient.
Biobank-scale datasets currently available 
consist of hundreds of thousands of whole genomes 
and hundreds of terabytes of compressed VCF.
Row-based data storage is fundamentally unsuited to such datasets
and these large VCFs represent a major computational bottleneck.

@@

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, storing information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank-scale projects such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.

VCF's simple text-based design and widespread support makes it an 
excellent archival format, but it is a poor basis for computational
workflows. Methods that require efficient access to genotype data
either require convertion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use 
bespoke binary formats~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate} 
that support the required access patterns.
While PLINK and BGEN support efficient access to genotype data, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy. PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format also leads to large files, particularly for recent
whole-genome sequence datasets. For example, at two bits 
per genotype, the full matrix for the GraphTyper SNP dataset in the UKB
WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

Downstream processing of these datasets can be split into a 
few broad categories. The first level of analysis that is 
done is quality control (QC). Variant calling QC is an 
involved and multi-faceted 
task [CITATIONS?], requiring analysis of many different metrics
and properties of the data and substantial computation.
Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
for quality control of large-scale variation datasets.
Building on components from the Hadoop ecosytem~\citep{white2012hadoop},
Hail distributes calculations and data transparently over a cluster, enabling
interactive analysis of very large datasets. It has been successfully used 
in many studies, and instrumental in datasets such as 
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
After QC, a common task is to perform a genome wide association study
(GWAS)~\cite{uffelmann2021genome}. The majority of tools for 
performing GWAS and associated tasks require
data to be in PLINK format~\cite[e.g][]{chang2015second} [MORE],
and so data must be ``hard-called'' according to some QC criteria
and exported to additional (large) files.
Many applications examine data on a variant-by-variant
basis, but others,
e.g.\ reference-panel based phasing and imputation 
methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
and Ancestral Recombination Graph 
(ARG) inference~\cite{brandt2024promise,lewanski2024era,wong2024general}
require access on a sample-by-sample basis.
Finally, variation datasets are often queried in exploratory
analyses, to find regions or variants of interest for a particular 
study (see section XXX).

While Hail is powerful software, it is not a suitable platform for 
developing inference or performing complex analysis. It is complex,
vertically integrated software, with an undocumented file format
[say this better]. 
Similarly, commerical solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~[Terra? BigQuery?]
or single-vendor technologies~\cite[e.g.][]{tiledb2024}.
